---
title: "Mini Data-Analysis Deliverable 1"
output: github_document
---

# Welcome to your (maybe) first-ever data analysis project!

And hopefully the first of many. Let's get started:

1.  Install the [`datateachr`](https://github.com/UBC-MDS/datateachr) package by typing the following into your **R terminal**:

<!-- -->

    install.packages("devtools")
    devtools::install_github("UBC-MDS/datateachr")

2.  Load the packages below.

```{r}
library(datateachr)
library(tidyverse)
```

3.  Make a repository in the <https://github.com/stat545ubc-2022> Organization. You will be working with this repository for the entire data analysis project. You can either make it public, or make it private and add the TA's and Lucy as collaborators. A link to help you create a private repository is available on the #collaborative-project Slack channel. 

# Instructions

## For Both Milestones

-   Each milestone is worth 45 points. The number of points allocated to each task will be annotated within each deliverable. Tasks that are more challenging will often be allocated more points.

-   10 points will be allocated to the reproducibility, cleanliness, and coherence of the overall analysis. While the two milestones will be submitted as independent deliverables, the analysis itself is a continuum - think of it as two chapters to a story. Each chapter, or in this case, portion of your analysis, should be easily followed through by someone unfamiliar with the content. [Here](https://swcarpentry.github.io/r-novice-inflammation/06-best-practices-R/) is a good resource for what constitutes "good code". Learning good coding practices early in your career will save you hassle later on!

## For Milestone 1

**To complete this milestone**, edit [this very `.Rmd` file](https://raw.githubusercontent.com/UBC-STAT/stat545.stat.ubc.ca/master/content/mini-project/mini-project-1.Rmd) directly. Fill in the sections that are tagged with `<!--- start your work below --->`.

**To submit this milestone**, make sure to knit this `.Rmd` file to an `.md` file by changing the YAML output settings from `output: html_document` to `output: github_document`. Commit and push all of your work to the mini-analysis GitHub repository you made earlier, and tag a release on GitHub. Then, submit a link to your tagged release on canvas.

**Points**: This milestone is worth 45 points: 43 for your analysis, 1 point for having your Milestone 1 document knit error-free, and 1 point for tagging your release on Github.

# Learning Objectives

By the end of this milestone, you should:

-   Become familiar with your dataset of choosing
-   Select 4 questions that you would like to answer with your data
-   Generate a reproducible and clear report using R Markdown
-   Become familiar with manipulating and summarizing your data in tibbles using `dplyr`, with a research question in mind.

# Task 1: Choose your favorite dataset (10 points)

The `datateachr` package by Hayley Boyce and Jordan Bourak currently composed of 7 semi-tidy datasets for educational purposes. Here is a brief description of each dataset:

-   *apt_buildings*: Acquired courtesy of The City of Toronto's Open Data Portal. It currently has 3455 rows and 37 columns.

-   *building_permits*: Acquired courtesy of The City of Vancouver's Open Data Portal. It currently has 20680 rows and 14 columns.

-   *cancer_sample*: Acquired courtesy of UCI Machine Learning Repository. It currently has 569 rows and 32 columns.

-   *flow_sample*: Acquired courtesy of The Government of Canada's Historical Hydrometric Database. It currently has 218 rows and 7 columns.

-   *parking_meters*: Acquired courtesy of The City of Vancouver's Open Data Portal. It currently has 10032 rows and 22 columns.

-   *steam_games*: Acquired courtesy of Kaggle. It currently has 40833 rows and 21 columns.

-   *vancouver_trees*: Acquired courtesy of The City of Vancouver's Open Data Portal. It currently has 146611 rows and 20 columns.

**Things to keep in mind**

-   We hope that this project will serve as practice for carrying our your own *independent* data analysis. Remember to comment your code, be explicit about what you are doing, and write notes in this markdown document when you feel that context is required. As you advance in the project, prompts and hints to do this will be diminished - it'll be up to you!

-   Before choosing a dataset, you should always keep in mind **your goal**, or in other ways, *what you wish to achieve with this data*. This mini data-analysis project focuses on *data wrangling*, *tidying*, and *visualization*. In short, it's a way for you to get your feet wet with exploring data on your own.

And that is exactly the first thing that you will do!

1.1 Out of the 7 datasets available in the `datateachr` package, choose **4** that appeal to you based on their description. Write your choices below:

**Note**: We encourage you to use the ones in the `datateachr` package, but if you have a dataset that you'd really like to use, you can include it here. But, please check with a member of the teaching team to see whether the dataset is of appropriate complexity. Also, include a **brief** description of the dataset here to help the teaching team understand your data.

<!-------------------------- Start your work below ---------------------------->

1: apt_buildings

2: building_permits

3: parking_meters

4: vancouver_trees

<!----------------------------------------------------------------------------->

1.2 One way to narrowing down your selection is to *explore* the datasets. Use your knowledge of dplyr to find out at least *3* attributes about each of these datasets (an attribute is something such as number of rows, variables, class type...). The goal here is to have an idea of *what the data looks like*.

*Hint:* This is one of those times when you should think about the cleanliness of your analysis. I added a single code chunk for you below, but do you want to use more than one? Would you like to write more comments outside of the code chunk?

<!-------------------------- Start your work below ---------------------------->

```{r}
### EXPLORE HERE ###
glimpse(apt_buildings) # returns dimension and every columns of a dataset
```
1. After running glimpse(apt_buildings), I know the number of rows in apt_buildings is 3455. I also know the number of cols in apt_buildings is 37. Meanwhile, I know the name and type of each variable in the apt_buildings dataset.

```{r}
### EXPLORE HERE ###
head(building_permits) # returns the first 6 rows of building_permits
dim(building_permits) # returns the dimension of building_permits
```
2. After running head(building_permits), I know the name and type of each variable in building_permits and get the first 6 rows of building_permits. After running dim(building_permits), I know the number of rows in building_permits is 20680 and the number of cols in building_permits is 14.

```{r}
### EXPLORE HERE ###
class(parking_meters) # returns the overall structure of the data
head(parking_meters) # returns the first 6 rows of parking_meters
```
3. After running class(parking_meters), I know the class of parking_meters is data.frame and "tbl" and "tbl_df" are a modern taken on the parking_meters data. After running head(parking_meters), I know the name and type of each variable in parking_meters and get the first 6 rows of parking_meters. 

```{r}
### EXPLORE HERE ###
head(vancouver_trees) # returns the first 6 rows of vancouver_trees
summary(vancouver_trees) # returns summary statistics of vancouver_trees
```
4. After running head(vancouver_trees), I know the name and type of each variable in vancouver_trees and get the first 6 rows of vancouver_trees. After running summary(vancouver_trees), I know the min, 1st Qu, Median, Mean, 3rd Qu, and max statistics of each numerical variable in vancouver_trees dataset.

<!----------------------------------------------------------------------------->

1.3 Now that you've explored the 4 datasets that you were initially most interested in, let's narrow it down to 2. What lead you to choose these 2? Briefly explain your choices below, and feel free to include any code in your explanation.

<!-------------------------- Start your work below ---------------------------->

I will choose **apt_buildings** and **vancouver_trees** datasets. For apt_buildings dataset, it has both rich categorical and numerical variables. Besides, the variables of apt_buildings is easier to understand for me. Hence, I think I can construct a variety of research questions based on its diverse data. For vancouver_trees dataset, it has 146611 rows that is a really large dataset and I think a larger amount of data would give a more accurate answer to the question. Furthermore, I'm currently live in Vancouver, so I'm more interested in the dataset that is close to my life.

<!----------------------------------------------------------------------------->

1.4 Time for the final decision! Going back to the beginning, it's important to have an *end goal* in mind. For example, if I had chosen the `titanic` dataset for my project, I might've wanted to explore the relationship between survival and other variables. Try to think of 1 research question that you would want to answer with each dataset. Note them down below, and make your final choice based on what seems more interesting to you!

<!-------------------------- Start your work below ---------------------------->

For apt_buildings dataset, my research question is whether the type of heating in the apartment changes with the year the house was built?

For vancouver_trees dataset, my research question is what is the relationship between the height of tree and the year it was planted? Will a tree with a bigger year be higher?

My final decision is **vancouver_trees** dataset, because I'm more interested in the situation of tree in Vancouver.
<!----------------------------------------------------------------------------->

# Important note

Read Tasks 2 and 3 *fully* before starting to complete either of them. Probably also a good point to grab a coffee to get ready for the fun part!

This project is semi-guided, but meant to be *independent*. For this reason, you will complete tasks 2 and 3 below (under the **START HERE** mark) as if you were writing your own exploratory data analysis report, and this guidance never existed! Feel free to add a brief introduction section to your project, format the document with markdown syntax as you deem appropriate, and structure the analysis as you deem appropriate. Remember, marks will be awarded for completion of the 4 tasks, but 10 points of the whole project are allocated to a reproducible and clean analysis. If you feel lost, you can find a sample data analysis [here](https://www.kaggle.com/headsortails/tidy-titarnic) to have a better idea. However, bear in mind that it is **just an example** and you will not be required to have that level of complexity in your project.

# Task 2: Exploring your dataset (15 points)

If we rewind and go back to the learning objectives, you'll see that by the end of this deliverable, you should have formulated *4* research questions about your data that you may want to answer during your project. However, it may be handy to do some more exploration on your dataset of choice before creating these questions - by looking at the data, you may get more ideas. **Before you start this task, read all instructions carefully until you reach START HERE under Task 3**.

2.1 Complete *4 out of the following 8 exercises* to dive deeper into your data. All datasets are different and therefore, not all of these tasks may make sense for your data - which is why you should only answer *4*. Use *dplyr* and *ggplot*.

1.  Plot the distribution of a numeric variable.
2.  Create a new variable based on other variables in your data (only if it makes sense)
3.  Investigate how many missing values there are per variable. Can you find a way to plot this?
4.  Explore the relationship between 2 variables in a plot.
5.  Filter observations in your data according to your own criteria. Think of what you'd like to explore - again, if this was the `titanic` dataset, I may want to narrow my search down to passengers born in a particular year...
6.  Use a boxplot to look at the frequency of different observations within a single variable. You can do this for more than one variable if you wish!
7.  Make a new tibble with a subset of your data, with variables and observations that you are interested in exploring.
8.  Use a density plot to explore any of your variables (that are suitable for this type of plot).

<!-------------------------- Start your work below ---------------------------->

1. Plot the distribution of the number of trees over years with histogram. (Finish exercise 1)
```{r}
tree_year_distribution <- ggplot(vancouver_trees, aes(date_planted)) + 
  geom_histogram(bins = 40) # use 40 bins in histogram
print(tree_year_distribution)
```

2. Filter observations of trees that have diameter wider than 9 to get trees wider than half of all trees. (Finish exercise 5)
```{r}
# 9 is the median diameter
wide_vancouver_trees <- filter(vancouver_trees, diameter > 9) %>% # filter all trees that have diameter wider than 9
  select(diameter, everything()) %>% # move diameter variable as the first column
  arrange(diameter) # sort the rows by diameter in ascending order
head(wide_vancouver_trees) # print the first 6 rows of wide_vancouver_trees
```

3.  Make a new tibble with tree_id, genus_name, height_range_id, diameter and date_planted. (Finish exercise 7)
```{r}
vancouver_trees_genus <- vancouver_trees %>%
  select(tree_id, genus_name, height_range_id, diameter, date_planted) %>% # select cols that I'm interested in
  arrange(genus_name) %>% # arrange the rows by genus_name
  drop_na(genus_name) # drop NA value in genus_name
head(vancouver_trees_genus) # print the first 6 rows of vancouver_trees_genus
```

4.  Use a density plot to explore the distribution of genus of trees over years. (Finish exercise 8)
```{r}
genus_distribution1 <- vancouver_trees_genus %>% 
  drop_na(date_planted) %>% # drop NA values in date_planted
  ggplot(aes(x = date_planted)) + 
  geom_density(aes(color = genus_name, fill = genus_name, alpha = 0.1)) # draw density of each genus in one plot
print(genus_distribution1)
```
```{r}
# Another way to draw the distribution of genus of trees over years
genus_distribution2 <- vancouver_trees_genus %>%
  drop_na(date_planted) %>%
  ggplot(aes(date_planted, genus_name)) + 
  ggridges::geom_density_ridges(bandwidth = 500) # parallel draw density of each genus 
print(genus_distribution2)
```

<!----------------------------------------------------------------------------->

2.2 For each of the 4 exercises that you complete, provide a *brief explanation* of why you chose that exercise in relation to your data (in other words, why does it make sense to do that?), and sufficient comments for a reader to understand your reasoning and code.

<!-------------------------- Start your work below ---------------------------->

1. I am curious about the change in the number of trees planted each year. The distribution of number of trees over years can answer my question, which is related to the exercise 1. Histogram is useful to show the distribution of a continuous variable, so I choose it.

2. From the summary statistics of diameter, I know the median diameter of all trees is 9. I'm interested in the trees that have diameter wider than 9, so they are wider than half of all trees. By filtering the trees that have diameter than 9 out, I can get my answer. Hence, it is related to exercise 5.

3. I'm interested in tree_id, genus_name, height_range_id, diameter and date_planted variables, so I select them and created a new tibble with those variables that related to exercise 7.

4. I want to know how the number of trees planted in each genus varies with the year. Since the distribution of genus of trees can answer my question, I use the density plot to show the distribution, which is related to exercise 8. Besides, there are a lot kind of genus for trees, so I use two different density plot to find which shows better result.

<!----------------------------------------------------------------------------->

# Task 3: Write your research questions (5 points)

So far, you have chosen a dataset and gotten familiar with it through exploring the data. Now it's time to figure out 4 research questions that you would like to answer with your data! Write the 4 questions and any additional comments at the end of this deliverable. These questions are not necessarily set in stone - TAs will review them and give you feedback; therefore, you may choose to pursue them as they are for the rest of the project, or make modifications!

<!--- *****START HERE***** --->

1. Which genus of trees are most popular and which are planted the least in Vancouver?
2. Which decade saw more trees planted in Vancouver?
3. What is the relationship between the height of a tree and its age? Are older trees generally higher?
4. Which genus of trees are more likely to be thicker?

# Task 4: Process and summarize your data (13 points)

From Task 2, you should have an idea of the basic structure of your dataset (e.g. number of rows and columns, class types, etc.). Here, we will start investigating your data more in-depth using various data manipulation functions.

### 1.1 (10 points)

Now, for each of your four research questions, choose one task from options 1-4 (summarizing), and one other task from 4-8 (graphing). You should have 2 tasks done for each research question (8 total). Make sure it makes sense to do them! (e.g. don't use a numerical variables for a task that needs a categorical variable.). Comment on why each task helps (or doesn't!) answer the corresponding research question.

Ensure that the output of each operation is printed!

**Summarizing:**

1.  Compute the *range*, *mean*, and *two other summary statistics* of **one numerical variable** across the groups of **one categorical variable** from your data.
2.  Compute the number of observations for at least one of your categorical variables. Do not use the function `table()`!
3.  Create a categorical variable with 3 or more groups from an existing numerical variable. You can use this new variable in the other tasks! *An example: age in years into "child, teen, adult, senior".*
4.  Based on two categorical variables, calculate two summary statistics of your choosing.

**Graphing:**

5.  Create a graph out of summarized variables that has at least two geom layers.
6.  Create a graph of your choosing, make one of the axes logarithmic, and format the axes labels so that they are "pretty" or easier to read.
7.  Make a graph where it makes sense to customize the alpha transparency.
8.  Create 3 histograms out of summarized variables, with each histogram having different sized bins. Pick the "best" one and explain why it is the best.

Make sure it's clear what research question you are doing each operation for!

<!------------------------- Start your work below ----------------------------->

1. Which genus of trees are most popular and which are planted the least in Vancouver?
```{r}
# Implement 2.
# Computer the number of observations of each genus in genus_name (categorical variable).
genus_count <- vancouver_trees %>%
  select(genus_name) %>% # select only genus_name from vancouver_trees dataset
  group_by(genus_name) %>% # group rows by genus_name
  mutate(count = n()) %>% # add a new column to count the number of each genus
  arrange(count) %>% # arrange rows by count in ascending order
  distinct() # delete duplicated rows
genus_count
```

It helps answer the question. The ACER (36062 is the largest) is the most popular and KALOPANAX, HALESIA and ALBIZIA are planted the least (1 is the least).

```{r}
# Implement 5.
# Use a bar plot to show number of each genus and use vertical lines to help determine the accurate number.
genus_bar <- ggplot(vancouver_trees, aes(x = genus_name)) + 
  geom_bar(color = "black", fill = "white") + # geom_bar layer
  geom_hline(yintercept = seq(from = 0, to = 36000, by = 3000), color = "blue", linetype = "dashed", size = 1) + # geom_hline layer
  coord_flip() # flip the plot
print(genus_bar)
```

It helps answer the question. The ACER has the longest bar, which mean it's most popular. The KALOPANAX, HALESIA and ALBIZIA have the shortest bar, which means they are planted the least.

2. Which decade saw more trees planted in Vancouver?
```{r}
# Implement 3.
# The range of year in vancouver_trees is from 1989 to 2019. Create year_period (new categorical variable) as a new col that contains 3 decades with date_planted variable (existing numerical variable).
# Use height_years_trees to help count number of trees in each decade. 
height_years_trees <- vancouver_trees %>% 
  select(height_range_id, date_planted) %>% # select height_range_id and date_planted variables from vancouver_trees
  # create a new column to count number of trees that planted in each decade
  # assign 1989~1999 label to trees planed in this range
  # assign 2000~2010 label to trees planed in this range
  # assign 2011~2019 label to trees planed in this range
  mutate(year_period = ifelse(date_planted <= '1999-12-31', "1989~1999", 
                       ifelse(date_planted >= '2000-1-1' & date_planted <= '2010-12-31', "2000~2010", 
                       ifelse(date_planted >= '2011-1-1', "2011~2019", NA))))
height_years_trees
```
```{r}
# Count the number of trees in each decade based on height_years_trees dataset.
decades_trees_count <- height_years_trees %>% 
  group_by(year_period) %>% # group rows by each decade
  mutate(count=n()) %>% # create new column to store the number of trees in each decade 
  arrange(count) %>% # arrange the rows by count in ascending order
  distinct(year_period, count) # delete duplicated rows
decades_trees_count
```

It helps answer the question. The decade with the most number of trees planted is 2000~2010. 

```{r}
# Implement 8.
# Create 3 histograms and select different bins for them to find the best size.
# The 3 bins histogram can show the distribution of number of trees over years.
# 3 bins is the best option for answering the question, because we have 3 decades and it's easier to see which decade planted the most number of trees in 3 bins.
tree_distribution_bin1 <- ggplot(vancouver_trees, aes(date_planted)) +
  geom_histogram(bins = 3) + # 3 bins
  labs(x = "Year of planting", y = "Number of trees", title = "bins = 3") # add titles to the plot
print(tree_distribution_bin1)
```
```{r}
# Implement 8.
# The 10 bins histogram can show the distribution of number of trees over years.
# It's harder to compare the number of trees in each decade with 10 bins.
tree_distribution_bin2 <- ggplot(vancouver_trees, aes(date_planted)) +
  geom_histogram(bins = 10) + # 10 bins
  labs(x = "Year of planting", y = "Number of trees", title = "bins = 10") # add titles to the plot
print(tree_distribution_bin2)
```
```{r}
# Implement 8.
# The 50 bins histogram can show the distribution of number of trees over years.
# It's hardest to compare the number of trees in each decade with 50 bins.
tree_distribution_bin3 <- ggplot(vancouver_trees, aes(date_planted)) +
  geom_histogram(bins = 50) + # 50 bins
  labs(x = "Year of planting", y = "Number of trees", title = "bins = 50") # add titles to the plot
print(tree_distribution_bin3)
```

It helps answer the question. From histogram with 3 bins, we can easily see the decade with the most number of trees planted is 2000~2010.

3. What is the relationship between the height of a tree and its age? Are older trees generally higher?
```{r}
# Implement 2.
# Compute the number of observations of each height id and each decade in height_range_id (categorical variable) and year_period (categorical variable). In each height range id, compare the number of trees in each decade.
height_trees_count <- height_years_trees %>% 
  group_by(height_range_id, year_period) %>% # group rows by height_range_id and year_period
  mutate(height_count = n()) %>% # create a new column to store the number of trees in each height range and decade
  distinct(height_range_id, year_period, height_count) %>% # delete duplicated rows
  arrange(height_range_id, year_period) # arrange rows by height_range_id and year_period in ascending order
height_trees_count
```

It helps answer the question. From height_range_id 2 to 6, trees planted in 1989~1999 are more than 2000~2010 in each height_range_id. Also, trees planted in 2000~2010 are more than 2011~2019 in each height_range_id. From last question, we know the most number of trees planted was in 2000~2010, so it's more likely the height of the tree was roughly positively correlated with its age. The older trees are more likely to be higher than younger trees.

```{r}
# Implement 7.
# Use point plot and set its alpha transparency to 0.1 to show the relationship between the height of a tree and its age.
height_pointplot <- ggplot(vancouver_trees) +
  geom_point(aes(x = date_planted, y = height_range_id), alpha = 0.1) # point plot and set alpha transparency to 0.1
print(height_pointplot)
```

It helps answer the question. From point plot, we can see there are more points in earlier years that means more higher trees are planted in early years. Therefore, the height of the tree was roughly positively correlated with its age.

4. Which genus of trees are more likely to be thicker?
```{r}
# Implement 1.
# Use summary statistics of diameter to help determine which genus of trees are more likely to be thicker.
diameter_summary <- vancouver_trees %>%
  group_by(genus_name) %>% # group rows by genus_name
  drop_na(diameter) %>% # drop NA values
  # calculate min, Q1, median, mean, Q2, max and range of diameter
  summarise(min = min(diameter), Q1 = quantile(diameter, probs = 1/4), median = median(diameter), mean = mean(diameter), Q2 = quantile(diameter, probs = 3/4), max = max(diameter), range = max - mean)
diameter_summary
```

It helps answer the question. From the summary statistics, especially the max term can be seen that some genus tend to have wider diameter like ACER, RHAMNUS, and STYRAX.	

```{r}
# Implement 7.
# Use point plot and set its alpha transparency to 0.2 to show which genus of trees are more likely to be thicker.
diameter_pointplot <- ggplot(vancouver_trees) +
  geom_point(aes(x = diameter, y = genus_name), alpha = 0.2) # point plot and set alpha transparency to 0.2
print(diameter_pointplot)
```

It helps answer the question. There are only some points show really further to 0 diameter, which is ACER, RHAMNUS, and STYRAX.

<!----------------------------------------------------------------------------->

### 1.2 (3 points)

Based on the operations that you've completed, how much closer are you to answering your research questions? Think about what aspects of your research questions remain unclear. Can your research questions be refined, now that you've investigated your data a bit more? Which research questions are yielding interesting results?

<!-------------------------- Start your work below ---------------------------->

I can basically answer my research questions based on my operations. 

1. From the number of observations of each genus in genus_name and the bar plot, the KALOPANAX, HALESIA and ALBIZIA are planted the least and ACER is the most popular.

2. Based on the height_years_trees dataset that I created from vancouver_trees, I can count the number of trees in each decade. From the count and the histogram, we can find they planted the most number of trees in 2000~2010.

3. Trees planted in 1989~1999 usually have more higher trees than 2000~2010 and trees planted in 2000~2010 usually have more higher trees than 2011~2019. Since we know there are more trees in 2000~2010 than 1989~1999, it's more likely that the height of trees is positively related to its age. From the point plot, we can get the same answer.

4. From the summary statistics of diameter, the max diameter of each genus can especially show ACER, RHAMNUS, and STYRAX tend to have wider diameter than others. The point plot gives the same answer.

Aspects of research questions remain unclear:

Since there are a lot NA values in the vancouver_tree dataset, the answer I got based on the dataset could be wrong. 

Improvement:

I think we can think of some research questions that the answers won't be changed before and after deleting the NA values.

The most interesting result:

I think the answer to the 3rd research question validates my idea that older trees are usually higher than younger trees, which is the most interesting result to me.

<!----------------------------------------------------------------------------->

### Attribution

Thanks to Icíar Fernández Boyano for mostly putting this together, and Vincenzo Coia for launching.